Automatic Matching of Bullet Lands

### 2015-12-27 imaging-paper.Rnw
ratios, raw figure

% "Title of the paper"
\title{Automatic Matching of Bullet Lands}\thanksref{T1}
\thankstext{T1}{Resources for this manuscript, including R code to reproduce any of the models, tables and figures, can be found at  \url{https://github.com/erichare/imaging-paper}.}
\runtitle{Automatic Matching of Bullet Lands}

@@

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}

\runauthor{E. Hare et al.}
\end{aug}
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is the primary goal. We propose a framework for the analysis of the 3D surface measurements which transcribes the markings into a 2D plotting framework. This makes identification of matches easier and more rooted in science for both the expert and the lay person. The automatic matching routine we propose \hh{manages to (a) correctly identify lands with too much damage to be suitable for matching, and (b) correctly identify all 10,384 land-to-land matches.}
\end{abstract}

@@

\section{Introduction}
\hh{
Todo list for journal shape-up: there is a lot of good material, we mainly need to focus on what story to tell and make that a solid argument.
\begin{enumerate}
%\item take out a few things: no circular fit, no figures 9 and 10 
%\item do we need the web app?
%\item out: lots of the figures can go into a supplement, together with a discussion. In the paper we then have a summary of what can be seen, so that we can focus on the results. 
\item add: formula for loess fit and its s.e.
%\item add: comparison of error rates to those of FEs in Hamby study.
%\item change: emphasis in varying cross section. once a stable region is found, how little change is happening? 
\item add: Why do we take the lowest stable cross section?
\item add: more discussion of the density plots: probability ratio for a match
\item add: more discussion on additional bullet images. NIST database claims to have 70 bullets -- that would be two sets of the study, but only 35 bullets are currently available. Should we contact Alan Zheng to see whether we can get access to more bullet files?
\item add: should we add a paragraph on the speed of the evaluation? - separate out into pre-processing (needs to be done once) and matching (pairwise comparison of lands). If we only do bullet matching, we could speed this process up by eliminating some land comparisons.
\item definitions: 
A bullet's {\it signature} are the residual surface measurements after removing the big structure using the loess fit. The {\it cross section} defines the location along the $x$ axis  at which the signature is taken.
\end{enumerate}
}
Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

\hh{The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. }
\hh{XXX The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; however, no false-negative error rates for bullets nor details on land-to-land matches were given. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)
 }
\hh{XXX Matching of bullets using the BulletTrax3D system is described based on another set of ten consecutively rifled barrels in \citet{roberge:2006}. They describe a process using a bullet-to-bullet correlation score to identify matches. The authors state that this process `could be automated', but no implementation of the algorithm is available. XXX}

@@

\section{Evaluation}
\hh{In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
}
\hh{As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we assess cross-correlation of signatures at cross sections 25$\mu m$ apart. For a stable region, we require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands shows scratch marks across the surface, also known as `tank rash'.}
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Br6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Br9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unkn Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unkn Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}\hh{Images of the four lands that got flagged during the quality assessment. All of them show scratch marks across the striation marks from the barrel. They will be excluded from the remainder of the analysis.}}
\end{figure}
%
\hh{We exclude these four lands from the further matching considerations and 
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unkown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). }
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 known lands. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

@
\end{minipage}
%' \end{subfigure}
%' \begin{subfigure}[t]{\textwidth}\centering
%' \caption{\label{fig:cms-dens}Densities of maximum CMS for Known Matches (KM) and Known Non Matches (KNM).}
%' <<cms-densities, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='.6\\textwidth'>>=
%' x <- seq(0,29, by=0.03)
%' km <- sm.density(bstats$CMS[bstats$match],h=2, model="Normal", display="none", eval.points=x)
%' knm <- sm.density(bstats$CMS[!bstats$match],h=1, display="none", eval.points=x)
%' dframe <- data.frame(x = x, km=km$estimate, knm=knm$estimate)
%' dframe$kmtail <- rev(cumsum(rev(dframe$km)))/sum(dframe$km)
%' dframe$knmtail <- rev(cumsum(rev(dframe$knm)))/sum(dframe$knm)
%' 
%' densities <- data.frame(
%'   match = rep(c("KM", "KNM"), each=(length(x)+2)),
%'   x = rep(c(0, x, 29), times = 2),
%'   estimate = c(0, km$estimate, 0, 0, knm$estimate,0))
%' 
%' ggplot() +
%'   geom_polygon(aes(x=x, y=estimate, fill=match, group=match), colour="black",
%'                data=densities, alpha=0.5) +
%' xlab("maximum CMS") +
%'   scale_fill_brewer("", palette="Paired") +
%'   ylab("Density estimates") +
%'   theme_bw() + 
%'   theme(legend.position=c(1,1), legend.justification=c(1,1),
%'         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' #  geom_line(aes(x=x, y=pmin(0.2, kmtail/knmtail/10)), data=dframe) 
%' 
%' # ggplot(data=bstats) + 
%' #     geom_density(aes(x=cms, fill=match), alpha=0.5) +
%' #   xlab("maximum CMS") +
%' #   scale_fill_brewer("", palette="Paired") +
%' #   theme_bw() + 
%' #   theme(legend.position=c(1,1), legend.justification=c(1,1),
%' #         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' 
%' @
%' \end{subfigure}
\caption{\label{fig:cms}Distribution of maximal CMS (left). Conditional barchart \citep{hummel} on the right: heights show probability of match/non-match given a specific CMS. All land-to-land comparisons with at least 13 CMS are matches.}
\end{figure}
%
There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are \Sexpr{mean(bstats$cms[!bstats$match])} maximal CMS between known non-matches (with a standard deviation of \Sexpr{sd(bstats$cms[!bstats$match])}). Known matches share on average \Sexpr{mean(bstats$cms[bstats$match])} maximal CMS, with a standard deviation of \Sexpr{sd(bstats$cms[bstats$match])}. Secondly, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by \citet{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch, between two lands that share twelve consecutively matched striae. Visually we can easily tell that these two lands do not match well.

@@

>>>>

% "Title of the paper"
\title{Automatic Matching of Bullet Lands}\thanksref{T1}
\thankstext{T1}{Resources for this manuscript, including R code to reproduce any of the models, tables and figures, can be found at  \url{https://github.com/erichare/imaging-paper}.}
\runtitle{Automatic Matching of Bullet Lands}
\runauthor{E. Hare et al.}

@@

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is the primary goal. We propose a framework for the analysis of the 3D surface measurements which transcribes the markings into a 2D plotting framework. This makes identification of matches easier and more rooted in science for both the expert and the lay person. The automatic matching routine we propose \hh{manages to (a) correctly identify lands with too much damage to be suitable for matching, and (b) correctly identify all 10,384 land-to-land matches.}
\end{abstract}

@@

\section{Introduction}
\hh{
Todo list for journal shape-up: there is a lot of good material, we mainly need to focus on what story to tell and make that a solid argument.
\begin{enumerate}
%\item take out a few things: no circular fit, no figures 9 and 10 
%\item do we need the web app?
%\item out: lots of the figures can go into a supplement, together with a discussion. In the paper we then have a summary of what can be seen, so that we can focus on the results. 
%\item add: formula for loess fit and its s.e.
%\item add: comparison of error rates to those of FEs in Hamby study.
%\item change: emphasis in varying cross section. once a stable region is found, how little change is happening? 
\item add: Why do we take the lowest stable cross section?
\item add: more discussion of the density plots: probability ratio for a match
\item add: more discussion on additional bullet images. NIST database claims to have 70 bullets -- that would be two sets of the study, but only 35 bullets are currently available. Should we contact Alan Zheng to see whether we can get access to more bullet files?
\item add: should we add a paragraph on the speed of the evaluation? - separate out into pre-processing (needs to be done once) and matching (pairwise comparison of lands). If we only do bullet matching, we could speed this process up by eliminating some land comparisons.
\item definitions: 
A bullet's {\it signature} are the residual surface measurements after removing the big structure using the loess fit. The {\it cross section} defines the location along the $x$ axis  at which the signature is taken.
\end{enumerate}
}
Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

\hh{The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. }
\hh{The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)
 }
\hh{A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available. }

@@

\section{Evaluation}
\hh{In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
}
\hh{As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we assess cross-correlation of signatures at cross sections 25$\mu m$ apart. For a stable region, we require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands shows scratch marks across the surface, also known as `tank rash'.}
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Br6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Br9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unkn Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unkn Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}\hh{Images of the four lands that got flagged during the quality assessment. All of them show scratch marks across the striation marks from the barrel. They will be excluded from the remainder of the analysis.}}
\end{figure}
%
\hh{We exclude these four lands from the further matching considerations and 
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unkown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). }
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 known lands. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

@
\end{minipage}
%' \end{subfigure}
%' \begin{subfigure}[t]{\textwidth}\centering
%' \caption{\label{fig:cms-dens}Densities of maximum CMS for Known Matches (KM) and Known Non Matches (KNM).}
%' <<cms-densities, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='.6\\textwidth'>>=
%' x <- seq(0,29, by=0.03)
%' km <- sm.density(bstats$CMS[bstats$match],h=2, model="Normal", display="none", eval.points=x)
%' knm <- sm.density(bstats$CMS[!bstats$match],h=1, display="none", eval.points=x)
%' dframe <- data.frame(x = x, km=km$estimate, knm=knm$estimate)
%' dframe$kmtail <- rev(cumsum(rev(dframe$km)))/sum(dframe$km)
%' dframe$knmtail <- rev(cumsum(rev(dframe$knm)))/sum(dframe$knm)
%' 
%' densities <- data.frame(
%'   match = rep(c("KM", "KNM"), each=(length(x)+2)),
%'   x = rep(c(0, x, 29), times = 2),
%'   estimate = c(0, km$estimate, 0, 0, knm$estimate,0))
%' 
%' ggplot() +
%'   geom_polygon(aes(x=x, y=estimate, fill=match, group=match), colour="black",
%'                data=densities, alpha=0.5) +
%' xlab("maximum CMS") +
%'   scale_fill_brewer("", palette="Paired") +
%'   ylab("Density estimates") +
%'   theme_bw() + 
%'   theme(legend.position=c(1,1), legend.justification=c(1,1),
%'         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' #  geom_line(aes(x=x, y=pmin(0.2, kmtail/knmtail/10)), data=dframe) 
%' 
%' # ggplot(data=bstats) + 
%' #     geom_density(aes(x=cms, fill=match), alpha=0.5) +
%' #   xlab("maximum CMS") +
%' #   scale_fill_brewer("", palette="Paired") +
%' #   theme_bw() + 
%' #   theme(legend.position=c(1,1), legend.justification=c(1,1),
%' #         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' 
%' @
%' \end{subfigure}
\caption{\label{fig:cms}Distribution of maximal CMS (left). Conditional barchart \citep{hummel} on the right: heights show probability of match/non-match given a specific CMS. All land-to-land comparisons with at least 13 CMS are matches.}
\end{figure}
%
There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are \Sexpr{mean(bstats$CMS[!bstats$match], na.rm=T)} maximal CMS between known non-matches (with a standard deviation of \Sexpr{sd(bstats$CMS[!bstats$match])}). Known matches share on average \Sexpr{mean(bstats$CMS[bstats$match])} maximal CMS, with a standard deviation of \Sexpr{sd(bstats$CMS[bstats$match])}. While the probability for a match increases with the number of maximal CMS, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by \citet{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch, between two lands that share twelve consecutively matched striae. Visually we can easily tell that these two lands do not match well.

@@

<<prob.ratios, echo=FALSE>>=
library(reshape2)
dens.stats <- dm %>% group_by(var, variable) %>% mutate(
  sum = sum(value),
  up = cumsum(value)/sum,
  down = 1 - cumsum(value)/sum
)
ratios <- dens.stats %>% group_by(var, x) %>% summarise(
  upper = sum(up[variable=="KM"])/sum(up[variable=="KNM"]),
  lower = sum(down[variable=="KM"])/sum(down[variable=="KNM"])
)
rm <- melt(ratios, measure.var=c("upper", "lower"))
levels(rm$variable) <- c("P(X < c | KM) / P(X < c | KNM)", "P(X > c | KM) / P(X > c | KNM)")
qplot(x, value, colour=variable,  data=rm) + facet_wrap(~var, scales="free") + scale_y_log10() + theme_bw() + geom_hline(yintercept=1) + scale_colour_brewer(palette="Set2") + ylab("Probability Ratios")
@
